A Framework for Equivalence of Cast Calculi and a new Cast Representation

### 2020-09-30 paper.tex
edits to section 1

\todo[inline]{KC: Cite "Contracts for higher-order functions" for the proxy-based approach maybe?}

Gradual typing \citep{siek2006gradual,siek2015refined} includes both static typing 
and dynamic typing in the same language, enables smooth transition between them, while
enforcing type soundness.
To allow the smooth transition, gradually typed languages are featuerd by a special type
typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|, which enables 
imprecise type annotations. A gradual type checker allows programmers to cast 
implicitly between types that have no conflicting type information.
At runtime, casts store or check type information to enforce type soundness.
Researchers usually define the semantics of gradually typed languages by
translation to an intermediate language with explicit casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi \citep{siek2009exploring}.
% To allow the smooth transition, every gradually typed language
% has a special type typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|,
% which enables imprecise type annotations.
% Programmers can use an expression of type \texttt{T} in any context that \texttt{T} 
% is consistent to.
% Two types are consistent if they have no conflicting type information.
% For instance, \lstinline|(->  Bool Int)| is consistent with \lstinline|(->  $\star$ Int)| but
% not consistent with \lstinline|(->  Int Int)|.
% When the type of an expression is consistent but not equal to the type of its 
% context, there would be a runtime computation (i.e. a cast) that store or check 
% type information to enforce type soundness.

Unfortunately, straightforward implementations of casts impose significant 
runtime overheads that can change the asymptotic space complexity of a program 
\citep{herman2010space}.
For instance, the following module define functions 
that check the parity of natural numbers. The terms are ordinary, while the 
type annotations are not -- the output type of \lstinline|even?| is 
imprecise.
\begin{lstlisting}
(define (even? [n : Int]) : $\star$
  (if (= n 0) #t (odd? (- n 1))))
(define (odd? [n : Int]) : Bool
  (if (= n 0) #f (even? (- n 1))))
\end{lstlisting}
These functions look tail-recursive, so it is natural to expect that they 
consumes constant space. This is not the case, however, in a straightforward 
implementation because there is a cast around each of the 
``tail call'', one from \lstinline|Bool| to 
\lstinline|$\star$|, the other goes the other way.
A straightforward implementation would cast the output of the recursive 
call \textit{before} returning it, which effectively breaks the tail recursion 
and thus changes the space complexity.

Continuation is not the only place where casts might accumulate. 
Consider a variant of the above program written in continuation-passing style. 
Here, both functions returns \lstinline|Bool| so there is not pending cast in 
continuations. The continuation argument \lstinline|k|, however, is wrapped in 
a cast in each recursive call because the type of \lstinline|k| is less 
precise in \lstinline|even?|.
\begin{lstlisting}
(define (even? [n : Int] [k : (-> $\star$ Bool)]) : Bool
  (if (= n 0) (k #t) (odd? (- n 1) k)))
(define (odd? [n : Int] [k : (-> Bool Bool)]) : Bool
  (if (= n 0) (k #f) (even? (- n 1) k)))
\end{lstlisting}
Our examples programs are inspired by the examples in \citep{herman2010space}.
We direct interested readers to their work for more examples and detailed explanation.

A solution to this space-efficiency problem is representing cast information 
more concisely. There are several known space-efficient cast representations
\citep{Siek:2015:BCT:2737924.2737968,Siek:2010:TWB:1706299.1706342,Garcia:2013:CTB:2500365.2500603,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}.
The current state of the art includes
\begin{itemize}
	\item threesomes 
	\citep{Siek:2010:TWB:1706299.1706342,Garcia:2013:CTB:2500365.2500603},
	\item supercoercions \citep{Garcia:2013:CTB:2500365.2500603}, and
	\item coercions in normal form
	\citep{siek2012interpretations,Siek:2015:BCT:2737924.2737968}.
\end{itemize}

These cast representations are represented in their own dynamic semantics of a gradually typed language,
which make it harder to focus on key differences between these representations. 
Surprisingly, no one has published a dynamic semantics parameterized over cast representations.
We present in Section~\ref{sec:framework} first of such generic dynamic semantics in the form 
of a CEK abstract machine and formulate the interface between the machine and cast representations
as an abstract data type, namely Cast ADT.

Of course, an alternative cast representation must be proved
correct. That is, a language implementation should give the same output no 
matter whether it is using a cast representation to reduce space 
consumption. Our generic CEK machine comes with generic theorems (Section~\ref{sec:framework}),
with which researchers can prove a cast representation is correct by showing its corresponding
implementation of Cast ADT satisfies a bunch of properties. We believe this framework can reduce the burden
 of proving this kind of theorems and help people focusing on the cast representations.
For illustration, we apply the framework to Lazy UD coercions in normal form
and Lazy UD threesomes. 
We believe the framework is also applicable to other cast representations.
The framework and the illustration are available at the following URL:
\[\text{\urlFrameworkAndIllustration}\]

\todo[inline]{KC: (about the p below) Actually I think twosome's composition is structurally recursive but threesomes' is not}
The published cast representations have various strengths and weaknesses.
The major insight of these systems is a compose operator for casts which compress cast information to
reduce space consumption.
Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in proof assistants such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labeled bottom type), and
supercoercions (10 different kinds).

This paper presents a new cast representation 
(Section~\ref{sec:hypercoercions}), named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The straightforward Agda translation of composition 
operator for hypercoercions is able to pass the termination check and 
hypercoercions are suggestive of a bit-level representation that minimizes the 
need for pointers and fits all first-order casts into 64 bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: \LD\ and
\LUD~\citep{siek2009exploring}. With the \LD\ strategy, only 
\textbf{d}own 
casts (casts from \Tdyn\ to some other types) are subject to blame. 
With the 
\LUD\
strategy, however, some \textbf{u}p casts are also blamable.
We are interested in the D blame
tracking strategy because it comes with a more straightforward notion of
safe cast compared to \LUD~\citep{siek2009exploring}, which is why \LD\ 
was
chosen in the Grift compiler~\citep{kuhlenschmidt2018efficient}. We are also 
interested in \LUD\ because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The \LUD\ hypercoercions were inspired by the
supercoercions of \citep{Garcia:2013:CTB:2500365.2500603} (hence the name) and
the \LD\ hypercoercions were inspired by the normal forms of
\citep{siek2012interpretations}.
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper, we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that the eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

\section{Related Works}
\label{chap:related-works}

\todo[inline]{KC: how to make \textbackslash citet work?}

>>>>

\todo[inline]{KC: Cite "Contracts for higher-order functions" for the proxy-based approach maybe?}
\todo[inline]{JS: That paper doesn't exactly fit into what you're talking about in the above
sentence. It might be appopriate to cite it elsewhere in the paper. Also,
there are earlier papers about coercive subtyping (Computing with coercions by Breazu-Tannen et al)
that use a proxy-based approach. The contracts paper may be the first to propose blame tracking.}

Gradual typing \citep{siek2006gradual,siek2015refined} includes both static typing 
and dynamic typing in the same language and enables a smooth transition between them while
enforcing type soundness.
To enable a smooth transition, gradually typed languages feature a special type
typically written as \Tdyn{}, \lstinline|Dyn|, or \lstinline|Any|, which
provides a way to write imprecise type annotations. A gradual type checker allows programmers to cast 
implicitly between types that have no conflicting type information.
At runtime, casts record and check type information to enforce type soundness.
Researchers often define the semantics of gradually typed languages by
translation to an intermediate language with explicit casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi \citep{siek2009exploring}.
% To allow the smooth transition, every gradually typed language
% has a special type typically written as \Tdyn\ or \lstinline|Dyn| or \lstinline|Any|,
% which enables imprecise type annotations.
% Programmers can use an expression of type \texttt{T} in any context that \texttt{T} 
% is consistent to.
% Two types are consistent if they have no conflicting type information.
% For instance, \lstinline|(->  Bool Int)| is consistent with \lstinline|(->  $\star$ Int)| but
% not consistent with \lstinline|(->  Int Int)|.
% When the type of an expression is consistent but not equal to the type of its 
% context, there would be a runtime computation (i.e. a cast) that store or check 
% type information to enforce type soundness.

Unfortunately, straightforward implementations of casts impose significant 
runtime overheads that can change the asymptotic space complexity of a program 
\citep{herman2010space}.
For instance, consider the following functions 
that check the parity of natural numbers. The terms are ordinary, while the 
type annotations are not -- the output type of \lstinline|even?| is 
imprecise.
\begin{lstlisting}
(define (even? [n : Int]) : $\star$
  (if (= n 0) #t (odd? (- n 1))))
(define (odd? [n : Int]) : Bool
  (if (= n 0) #f (even? (- n 1))))
\end{lstlisting}
These functions look tail-recursive, so it is natural to expect that
they consumes constant space. This is not the case, however, in a
straightforward implementation because there is an implicit cast
around each of the recursive calls, one from \lstinline|Bool| to
\lstinline|$\star$| and the other from \lstinline|$\star$| to
\lstinline|Bool|.  A straightforward implementation would perform this
cast after the recursive calls but \textit{before} returning from the
current function. So the calls are not in fact tail calls and the
space complexity is linear instead of constant.

Continuations are not the only place where casts can accumulate.
Consider a variant of the above program written in
continuation-passing style.  Here, both functions returns
\lstinline|Bool| so there is no implicit cast in tail position.
However, the argument \lstinline|k| is wrapped in a cast in each
recursive call because the type of \lstinline|k| is less precise in
\lstinline|even?| than it is in \lstinline|odd?|.
\begin{lstlisting}
(define (even? [n : Int] [k : (-> $\star$ Bool)]) : Bool
  (if (= n 0) (k #t) (odd? (- n 1) k)))
(define (odd? [n : Int] [k : (-> Bool Bool)]) : Bool
  (if (= n 0) (k #f) (even? (- n 1) k)))
\end{lstlisting}
These example programs are adapted from those of Herman et al.~\citep{herman2010space}.
We direct readers to their work for more examples and a detailed explanation.

One solution to this space-efficiency problem is to represent cast
information more concisely. There are several known space-efficient
cast representations
\citep{Siek:2015:BCT:2737924.2737968,Siek:2010:TWB:1706299.1706342,Garcia:2013:CTB:2500365.2500603,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}.
The current state of the art includes
\begin{itemize}
	\item threesomes 
	\citep{Siek:2010:TWB:1706299.1706342,Garcia:2013:CTB:2500365.2500603},
	\item supercoercions \citep{Garcia:2013:CTB:2500365.2500603}, and
	\item coercions in normal form
	\citep{siek2012interpretations,Siek:2015:BCT:2737924.2737968}.
\end{itemize}

The semantics of these cast representations are defined in many
different papers with separate reduction rules, which makes it
difficult to isolate the key differences between them.  In this paper
we present the first set of reduction rules for a cast calculus that
is parameterized over the cast representation
(Section~\ref{sec:framework}). The semantics takes the form of a CEK
abstract machine and it formulates the interface between the machine
and the different cast representations as an abstract data type,
namely the \emph{Cast ADT}.

Of course, a space-efficient cast representation must be proved
correct. That is, a language implementation that uses space-efficient
casts should give the same output as the language semantics. Our
generic CEK machine comes with generic theorems
(Section~\ref{sec:framework}) that researchers can use to prove a cast
representation correct.  We believe this framework can reduce the
burden of proving correctness theorems and help people focus on the
cast representations.  For illustration, we apply the framework to
Lazy UD coercions in normal form~\citep{Siek:2015:BCT:2737924.2737968}
and Lazy UD threesomes~\citep{Siek:2010:TWB:1706299.1706342}. We
believe the framework is also applicable to other cast
representations.  The framework and illustrations are available at the
following URL:
\[\text{\urlFrameworkAndIllustration}\]

\todo[inline]{KC: (about the p below) Actually I think twosome's
  composition is structurally recursive but threesomes' is not}

The published space-efficient cast representations have various
strengths and weaknesses.  The major insight of these systems is a
compose operator for casts which compresses cast information to reduce
space consumption.  Threesomes and supercoercions are good for
mechanized metatheory because their compose operators are structurally
recursive, making them easy to define in proof assistants such as
Agda. In contrast, the coercions in normal form have compose operators
that are not structurally recursive, which makes it more difficult to
define in Agda, requiring what amounts to an explicit proof of
termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with their strange labeled bottom type), and
supercoercions (10 different kinds).

This paper presents a new cast representation named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation (Section~\ref{sec:hypercoercions}). The
straightforward Agda translation of the composition operator for
hypercoercions is able to pass Agda's termination check and
hypercoercions are suggestive of a bit-level representation that
minimizes the need for pointers and fits all first-order casts into 64
bits.
%
We present two flavors of hypercoercions to support the two blame
tracking strategies from the literature: \LD\ and
\LUD~\citep{siek2009exploring}. With the \LD\ strategy, only 
\textbf{d}own 
casts (casts from \Tdyn\ to some other types) are subject to blame. 
With the 
\LUD\
strategy, however, some \textbf{u}p casts may also be blamed.
We are interested in the D blame
tracking strategy because it comes with a more straightforward notion of
safe cast compared to \LUD~\citep{siek2009exploring}, which is why \LD\ 
was
chosen in the Grift compiler~\citep{kuhlenschmidt2018efficient}. We are also 
interested in \LUD\ because it plays a
prominent role in the gradual typing literature
\citep{wadler2009well}.  The \LUD\ hypercoercions were inspired by the
supercoercions of \citep{Garcia:2013:CTB:2500365.2500603} (hence the name) and
the \LD\ hypercoercions were inspired by the normal forms of
Siek and Garcia~\citep{siek2012interpretations}.
The semantics of casts can be lazy or
eager~\citep{siek2009exploring}. In this paper, we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because New et al.~\citep{new2019gradual} show that the eager
strategies are incompatible with $\eta$-equivalence of functions.

@@

\section{Related Work}
\label{chap:related-works}

\todo[inline]{KC: how to make \textbackslash citet work?}
\todo[inline]{JS: Good question! I would like to use citet as well.}
